---
title: "ENVS 193DS Final"
author: "Chloe Carlstrom"
date: "06-10-25"
toc: true
execute:
  warning: false
  message: false
format: html
---

Repository: https://github.com/chloecarlstrom/ENVS-193DS_spring-2025_final

# Reading in data/packages

```{r, warning=FALSE,message=FALSE}
library(tidyverse)
library(here)
library(janitor)
library(dplyr)
library(lubridate)
library(MuMIn)
library(DHARMa)
library(ggeffects)

sst <- read.csv(here("data", "SST_update2023.csv"))

nest_boxes <- read.csv(here("data", "occdist.csv"))

```

# Problem 1. Research Writing

## a. Transparent statistical methods
In part 1 they used a correlation test (such as Pearson's r) because they state state that they reject the null that there is no "correlation" between the two variables. In part 2 they likely used a one-way ANOVA because they are looking at average nitrogen load compared across source (factor with multiple levels). 

## b. More information needed
I think it would be useful to use a linear model alongside the correlation test to provide more information on the strength and direction of the correlation. For the ANOVA, it would be helpful to run a Tukey HSD to provide insight into which groups are actually different.

## c. Suggestions for rewriting
We found a strong positive relationship between distance from headwater (km) and annual total nitrogen load (kg year ^-1)(Pearson's r = correlation coefficient, p = 0.03). 

We found a large difference ($\eta^2 = 0.7$) in average nitrogen load (kg year^-1) between sources (urban land, atmospheric deposition, fertilizer, wastewater treatment, and grasslands) (one-way ANOVA, F(among groups DF, within groups DF) = F statistic, p =0.02, $\alpha = significance level).


# Problem 2. Data Visualization

## a. Cleaning and summarizing 
```{r, warning=FALSE, message=FALSE}
sst_clean <- sst %>% #naming new object from sst data frame
  mutate( #function to change data
    year = factor(year(date)), #making year as a factor from date column
    month = factor(month(date, label = TRUE), ordered = TRUE)) %>% #making a month factor from date
  group_by(year, month) %>% #grouping by year and month
  summarise(mean_monthly_sst = mean(temp, na.rm =TRUE), .groups = "drop") %>% #using month data to find mean monthly sst
  ungroup() %>% #ungrouping data
  select(year, month, mean_monthly_sst) %>% #showing only selected columns
  filter(year %in% c("2018", "2019", "2020", "2021", "2022", "2023")) 

slice_sample(sst_clean, n = 5)

str(sst_clean)


```

## b. Visualize the data
```{r, warning = FALSE, message=FALSE}
ggplot(data = sst_clean, aes(x = month, y = mean_monthly_sst, group = year, color = factor(year))) + #setting up data for figure, changing year to numeric
  geom_line() + #line geometry
  geom_point() + #point geometry
  scale_color_manual(values = c("2018" = "lightblue", "2019" = "deepskyblue", "2020" = "dodgerblue3", "2021" = "royalblue1", "2022" = "blue2", "2023" = "darkblue")) + #coloring lines in gradient of single color
  theme(
    panel.background = element_blank(), #removing background
    panel.grid = element_blank(), #removing grid lines
    panel.border = element_rect(color = "black", fill = NA), #creating line around plot
    axis.ticks = element_blank(), #removing axis ticks
    legend.position = c(0.1, 0.9), #adding legend position inside box
    legend.justification = c("left", "top") #formating legend position
  ) +
  labs( #naming axes, legend
    x = "Month", 
    y = "Mean Monthly sea surface temperature (C)", 
    color = "Year"
  )

```

# Problem 3. Data Analysis

## a. Response variable
The 0s and 1s are used to denote if the species is present in the nest box with 1 being yes and 0 being no. 

## b. Purpose of study
The Swift Parrots are critically endangered and the other two species, Tree Martins and Common Starlings are their only nest competitors. Both exploit nest boxes, with Tree Martins being a native species and Common Starlings being introduced. 

## c. Difference in "seasons"
The years represent breeding seasons when the parrots bred at the study site. These breeding events are triggered by mast tree flowering events. 

## d. Table of models
4 models total: 

| Model number |season| distance to forest edge |    description              |  
|:------------:|:---:|:------------------------:|-----------------------------|  
| 1            |     |                          | null model                  |
| 2            |  X  |           X              | saturated model             | 
| 3            |  X  |                          | season and sp occupancy     |   
| 4            |     |           X              | distance and sp occupancy   |    


## e. Run the models
```{r, warning=FALSE, message=FALSE}
nest_boxes <- nest_boxes %>% 
  mutate(
    season = factor(season) #making season into a factor
  )

str(nest_boxes) #showing structure to ensure variables have correct form

#null model
model1 <- glm(sp ~ 1, #predictor and response variables
              data = nest_boxes, #data frame
              family = "binomial") #data form

#saturated model
model2 <- glm(sp ~ season + edge.distance, #predictor and response variables
              data = nest_boxes, #data frame
              family = "binomial") #data form

#model with season only
model3 <- glm(sp ~ season, #predictor and response variables
              data = nest_boxes, #data frame
              family = "binomial") #data form

#model with distance to edge only
model4 <- glm(sp ~ edge.distance, #predictor and response variables
              data = nest_boxes, #data frame
              family = "binomial") #data form

```

## f. Check the diagnostics
```{r, warning=FALSE, message=FALSE}
#diagnostics for each model

#model 1
plot(
  simulateResiduals(model1)
     )

#model 2
plot(
  simulateResiduals(model2)
)

#model 3
plot(
  simulateResiduals(model3)
)

#model 4
plot(
  simulateResiduals(model4)
)


```

## g. Select the best model
```{r, warning=FALSE, message=FALSE}
AICc( #determining best model
  model1,
  model2,
  model3,
  model4
) %>% 
arrange(AICc)

```
The best model as determined by the Akaike's Information Criterion (AIC) was the saturated model with the combined effects of both season and distance from forest edge on nest box occupancy. 

## h. Visualize the model predictions
```{r, warning=FALSE, message=FALSE}
mod_preds <- ggpredict(model2, #predictions for model 2 using combined variables of edge distance and season
                       terms = c("edge.distance", "season"))
ggplot() +
  geom_point(data = nest_boxes, #adding data points
             aes(x = edge.distance, y = sp), #variables
             alpha = 0.3) + #size of points
  geom_ribbon(data = mod_preds, #adding CI ribbon
              aes(x = x, #variables
                  ymin = conf.low, 
                  ymax = conf.high, 
                  fill = group), #filling by year
              alpha = 0.2) +
  geom_line(data = mod_preds, #adding prediction line
            aes(x = x, y = predicted, color = group), size = 1) + 
  scale_y_continuous(limits = c(0, 1)) + #setting limits for y axis
  labs( #titles
    x = "Distance to Forest Edge",
    y = "Swift Parrot Nest Box Occupancy",
    color = "Season", 
    fill = "Season"
  ) +
  scale_fill_manual( #coloring ribbon
    values = c(
      "2016" = "slateblue1",
      "2019" = "mediumorchid3"
    )) +
  scale_color_manual( #coloring lines
    values = c(
      "2016" = "darkslateblue",
      "2019" = "darkmagenta"
    ))


```

**Figure 1. Predicted Swift Parrot nest box occupancy related distance to forest edge grouped by season** Swift parrots have a higher probability of occupying nest boxes closer to the forest edge. Overall the probability of nest boxes being occupied was higher 2016 than 2019. Stojanovic, Dejan et al. (2021). Do nest boxes breed the target species or its competitors? A case study of a critically endangered bird [Dataset]. Dryad. https://doi.org/10.5061/dryad.83bk3j9sb

## j. Calculate model predictions
```{r, warning=FALSE, message=FALSE}

ggpredict(model2, #model used
          terms = c("edge.distance [0]", "season")) #setting distance to 0, adding season

ggpredict(model2, #model used
          terms = c("edge.distance [900]", "season")) #setting distance to 900, adding season

```

## k. Interpret your results
For Swift Parrots, the probability of nest box occupancy decreases with increasing distance to forest edge. At forest edge (0m) the probability of occupancy was 0.48 in 2016 (95% CI: [0.33, 0.64]) and 0.30 in 2019 (95% CI: [0.18, 0.44]). At 900m distance from forest edge, the probability of occupancy was 0.12 in 2016 (95% CI: [0.06, 0.24]) and 0.06 in 2019 (95% CI:[0.03, 0.13]). This could be due to the fact that Tree Martins may have outcompeted Swift Parrots for next boxes farther from forest edge, forcing them to occupy nest boxes closer to forest edge. 

# Problem 4. Affective and exploratory visualizations

## a. Comparing visualizations

### How are the visualizations different from each other in the way you have represented your data?
My visualizations from homework 2 showcased the exact numerical data whereas my affective visualization showcased my data in a way that is more interesting to look and comprehend without fully displaying the underlying data. 

### What similarities do you see between all your visualizations?
The overall trends of my data and the way in which my data is organized, for example the axes generally had the same variables and were grouped in similar manners.

### What patterns (e.g. differences in means/counts/proportions/medians, trends through time, relationships between variables) do you see in each visualization? Are these different between visualizations? If so, why? If not, why not?
Across my visualizations, although there were no strong relationships, I noticed that overall I generally had lower step counts on days where I spent more time on campus. I also had higher step counts on days that I ran. I largely did longer workouts such as bike rides and hikes on days where I spent little to no time on campus. 

### What kinds of feedback did you get during week 9 in workshop or from the instructors? How did you implement or try those suggestions? If you tried and kept those suggestions, explain how and why; if not, explain why not.
I got feedback to overall enlarge my visualization as it was a bit hard to decipher. I also labeled the y-axis with hour marks so that the data was better explained with actual numbers. I also go feedback that it was a bit difficult to tell the different size of footprints so I tried to make the variation more obvious and in the key I labeled what each footprint size meant in terms of 1000s of steps. 

